---
title: "Telecom Customer Retention Project in R"
output: html_notebook
---
# Purpose : Analyze customer data for MTN Telecom in order to craft effective customer retention programs

# 1.Defining the Question

## a) Specifying the Data Analysis Question

The management would like to get your assistance in understanding the current subscribed customers. Your recommendations informed by your analysis will help them make informed decisions on effective customer retention programs.


## b) Defining the Metric for Success

The solution will have satisfied the research question when you can identify the factors that contribute to customer churn.

## c) Understanding the context

MTN Telecom offers mobile and internet services to its customers. These services include phone, multiple lines, internet, online security, online backup, device protection, tech support, and streaming TV and movies.

You have been provided with the current customer data. Since you will be working towards a descriptive report than a predictive one, you decide to think critically of the kind of questions that would help you craft customer retention programs. You then later use the given data set to answer your questions but before you start, you reading, explore, clean and visualise your dataset.

## d) Recording the Experimental Design

The steps taken included :

1.Data Loading & exploration 
2.Data cleaning.
4.Data analysis.
5.Data Visulaization -to effectively communicate your findings
5.Interpretation of results.
6.Recommendations based on results of the analysis.


## e) Data Relevance

For now, the data we have contains churn data which will be critical for our research specific analysis.

# 2. Data Cleaning & Preparation

```{r}
# loading libraries
library(tidyverse)

```

```{r}
# loading the CSV dataset from URL and loading the entire dataframe

customer_df <- read_csv("https://bit.ly/2ZlpzjF")
head(customer_df)

```


```{r}
# getting our dataset shape
#glimpse(customer_df)
#head(customer_df)
#tail(customer_df)
dim(customer_df)

```
This data set has 7050 records and 21 variables.

```{r}
# Checking the column names
names(customer_df)
```

  
```{r}
# Checking  duplicates if any are present

customer_df[duplicated(customer_df), ]

```

The data has 7 dublicated observations 

```{r}
# Remove duplicates and checking the shape of the resulting data

customer_df <- customer_df[!duplicated(customer_df), ]
dim(customer_df)
```
The no dublicate data now has unique  7043 observations and 21 varables after removing the dublicates 

```{r}
# Checking for duplicates if any are present after rmoving dublicates 

customer_df[duplicated(customer_df), ]

```
No dublicates found after removing the 7 observations


```{r}
#finding total missing values per column 
colSums(is.na(customer_df))
```
NB: Only customerID has no missing values , the rest have missing values from 1-23

```{r}
# dropping missing values

na.omit(customer_df)


```
40 missing records / values were dropped, which isn't a big loss.from 7043 to 7003 records 


```{r}
#No of missing values per column 

colSums(is.na(customer_df))

```
```{r}
names(customer_df)
```
Checking columns names 
Observation:
Names have mixed capital & lower case naming .

```{r}
# Change all column names to lower case

names(customer_df) <- tolower(names(customer_df))
names(customer_df)
```


```{r}
# previewing the number of unique values in each column

unique(customer_df)

```
 

```{r}
# checking the datatypes
str(customer_df)

```
```{r}
# checking unique values of each variable

customer_df2 <- customer_df %>% as.data.frame()

for (i in colnames(customer_df2)[-c(1, 6, 19, 20)]){
  
  cat("Unique values in", i, ":", unique(customer_df2[,i]), "\n")
}
```
There are columns with non unique values like payment method , senior citizen etc 

```{r}
# Replacing values in paymentmethod column

customer_df$paymentmethod[customer_df$paymentmethod == "Mailed checkkk"] <- "Mailed check"
customer_df$paymentmethod[customer_df$paymentmethod == "Electronic checkk"] <- "Electronic check"
```

```{r}
# resolving the issues in the 'payment_method' variable
unique(customer_df$paymentmethod)
```
```{r}
# Replacing values in seniorcitizen column

customer_df$seniorcitizen[customer_df$seniorcitizen == 0] <- 'No'
customer_df$seniorcitizen[customer_df$seniorcitizen == 1] <- 'Yes'
```

```{r}
# Recheck unique values
unique(customer_df$seniorcitizen)
```
```{r}
# Getting outliers based on totalcharges column
lower_bound <- quantile(customer_df$totalcharges, 0.025, na.rm=TRUE)
upper_bound <- quantile(customer_df$totalcharges, 0.975,na.rm=TRUE)

```

```{r}
#lower boound
lower_bound
```

```{r}
#upper boound
upper_bound
```

```{r}
# Getting the records without outlier values
non_outlier_values <- which(customer_df$totalcharges > lower_bound & customer_df$totalcharges < upper_bound)

```

```{r}
#non_outlier_values
customer_df = customer_df[non_outlier_values,]
dim(customer_df)

```
```{r}
# Remove duplicates

customer_df <- customer_df[!duplicated(customer_df), ]
dim(customer_df)
```


```{r}
# Remove all missing values

customer_df <- na.omit(customer_df)
dim(customer_df)
```
```{r}
#preview
head(customer_df)
customer_df
```

```{r}
#column names

names(customer_df)

```

```{r}
#load ggplot2 library 
library(ggplot2)
```

```{r}
#plot density distribution of churn wrt senior citizen 
ggplot(customer_df, aes(x = churn, fill = "churn")) +
  geom_density(alpha = 0.5) +
  geom_density(aes(x = seniorcitizen, fill = "seniorcitizen"), alpha = 0.5) +
  labs(title = "Density Plot of churn and seniorcitizen",
       x = "Value",
       y = "Density") +
  scale_fill_manual(values = c("churn" = "blue", "seniorcitizen" = "red"))


```


```{r}
# we investigate the distribution of churn by senior citizen
# and recording our observations.
# bar plot distribution of churn wrt senior citizen
ggplot(customer_df, aes(x = churn, fill = factor(seniorcitizen))) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by Senior Citizen",
       x = "Churn",
       y = "Count",
       fill = "Senior Citizen") +
  scale_fill_discrete(name = "Senior Citizen", labels = c("Non-Senior", "Senior")) +
  theme_minimal()
```



```{r}
# plotting a pie chart to investigate whether MTN customers churned
# Calculating churn counts
df <- table(customer_df$churn)
print(df)

```
```{r}
   pie(df )
  
```
```{r}
#What percentage of customers from the provided data set churned?
  churn_agg_tbl <- customer_df %>% group_by(churn) %>% 
 
   summarise( percent = 100 * n() / nrow(customer_df) )
   churn_agg_tbl
  
```
 
 Yes , 
 
 26% of the customers churned out from the customer data analysis .
 
 
  
```{r}
# let's now make a comparison betweeen male and female 
# subscribers by plotting the distribution of churn by gender

ggplot(customer_df, aes(x = gender, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by Gender",
       x = "Gender",
       y = "Count",
       fill = "Churn") +
  theme_minimal()
```

From the Analysis , there is near equal churn rate in Male and female showing churn rate is less /Not affected by gender.


```{r}
# Investigate the distribution of churn by dependents 
# and then record your observations.

# Filter out churned customers and compare numbers of customers with dependents vs those without

dependents_agg_tbl <- filter(customer_df, churn == 'Yes') %>% group_by(dependents) %>% 
   summarise(count=n(),
            .groups = 'drop')
dependents_agg_tbl

```
There is more churn out by customers without dependants than with dependants 


```{r}
#Investigate the distribution of churn by phone service:
# Filter out churned customers and compare numbers of customers with phone service vs those without

phone_agg_tbl <- filter(customer_df, churn == 'Yes') %>% group_by(phoneservice) %>% 
   summarise(count=n(),
            .groups = 'drop')

phone_agg_tbl

```
```{r}
#plotting Churn Distribution by phoneservice

plot <- ggplot(customer_df, aes(x = phoneservice, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by phoneservice",
       x = "phoneservice",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)
```

The rate of churn is higher among customers with phone service.

```{r}
# plotting the distribution of churn by multiple lines
# and record our observations.

plot <- ggplot(customer_df, aes(x = multiplelines, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by multiplelines",
       x = "multiplelines",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)

```
There is less churn by customers with NO phone service and similar churn out by people with multiple lines .

People with multiple lines are less likely to churn out




```{r}
# plotting the distribution of churn by internet service 
# and later record our observations.

plot <- ggplot(customer_df, aes(x = internetservice, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by internetservice",
       x = "internetservice",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

 # Display the plot
 print(plot)

```
Greatest churn out seen in customers with fiber optic conection 
Minimal churn out is seen in people without internet service  



```{r}
# Investigate the distribution of churn by online security and then record
# our observations below:

plot <- ggplot(customer_df, aes(x = onlinesecurity, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by onlinesecurity",
       x = "onlinesecurity",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

 # Display the plot
 print(plot)


```
there is more churn by customers who lack online security compared to those with online security .

People without internet service less likely to chrun out


```{r}
# plotting the distribution of churn by online backup

ggplot(customer_df, aes(x = onlinebackup, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by onlinebackup",
       x = "onlinebackup",
       y = "Count",
       fill = "Churn") +
  theme_minimal()
```


```{r}
# plotting the distribution of churn by tech support
ggplot(customer_df, aes(x = techsupport, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by techsupport",
       x = "techsupport",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

```


```{r}
# Investigate the distribution of churn by device protection
plot <- ggplot(customer_df, aes(x = deviceprotection, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by deviceprotection",
       x = "deviceprotection",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)

```
Looking at device protection , more peple without device protection are likely to churn out.


```{r}
# plotting the distribution of churn by streaming tv
# Filter out churned customers and compare numbers of customers with streaming tv vs those without

tv_agg_tbl <- filter(customer_df, churn == 'Yes') %>% group_by(streamingtv) %>% 
   summarise(count=n(),
            .groups = 'drop')

tv_agg_tbl
```
The rate of churn is comparable for customers with streaming tv and those without.

```{r}
# plotting the distribution of churn by streaming movies

plot <- ggplot(customer_df, aes(x = streamingmovies, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by Streaming Movies",
       x = "Stream Movies",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)

```
Churn is less/unlikely affected by availability of internet services and so retention offers should not be based on internet service 

Streaming movies somehow affects the churn out rate as about 40-50% of the customers churned out . something need be checked on data speeds which could potentially affected movie streaming 


```{r}
# Investigate the distribution of churn by contract

plot <- ggplot(customer_df, aes(x = contract, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by contract",
       x = "contract",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)

```
Short term contract customers (month-month) are likely to churn out . long term contract (2 yrs) are less likely to churn out .

Make more peple have longer contract to reduce churn 


```{r}
#Investigate the distribution of churn by partner:
# Filter out churned customers and compare numbers of partner vs non partner

partner_agg_tbl <- filter(customer_df, churn == 'Yes') %>% group_by(partner) %>% 
   summarise(count=n(),
            .groups = 'drop')
partner_agg_tbl
```

More churn noted in Non partners than with partner subs  


```{r}
# Investigate  the distribution of churn by paperless billing
#plotting the distribution of churn by paperless billing

plot <- ggplot(customer_df, aes(x = paperlessbilling, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by paperlessbilling",
       x = "paperlessbilling",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)


```
More churn is seen in customers with paperless billing .


```{r}
# plotting the distribution of churn by payment method

plot <- ggplot(customer_df, aes(x = paymentmethod, fill = churn)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Distribution by paymentmethod",
       x = "paymentmethod",
       y = "Count",
       fill = "Churn") +
  theme_minimal()

# Display the plot
print(plot)


```

Customers with electronic check  payment method are more likely to churn out.


#b) General Analysis.

## Comparing customers by gender:

```{r}
# compare customers in the data set by gender

gender_tbl <- customer_df %>% 
  group_by(gender) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
gender_tbl
```
the number of male and female are nearly equal

## Comparing senior vs non senior customers:

```{r}

# compare senior vs non-senior customers

senior_tbl <- customer_df %>% group_by(seniorcitizen) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
senior_tbl
```

Very few of the customers are senior citizens

## Compare customers with dependents and those without

```{r}
dependents_tbl <- customer_df %>% group_by(dependents) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
dependents_tbl
```
most of the customers have no dependants 

## Comparing customers by tenure

```{r}
tenure_tbl <- customer_df %>% group_by(tenure) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
tenure_tbl
```
Most customers have a tenure of below 10


## Comparing customers by phone service

```{r}
phone_tbl <- customer_df %>% group_by(phoneservice) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
phone_tbl
```
Most of the customers have phone service 

## Comparing customers by whether the have multiple lines:

```{r}
multiplelines_tbl <- customer_df %>% group_by(multiplelines) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
multiplelines_tbl
```

Customer with and and without multiple lines are nearly the same 

## Comparing customers by internet service:

```{r}
# distribution by internet service

internet_tbl <- customer_df %>% group_by(internetservice) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
internet_tbl
```
Most  of the  customers have internet service

##omparing customers by online security

```{r}
# distribution by online security

security_tbl <- customer_df %>% group_by(onlinesecurity) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
security_tbl
```

most customers lack online security

## Comparing customers by online backup

```{r}
# distribution by online backup

backup_tbl <- customer_df %>% group_by(onlinebackup) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
backup_tbl
```
Most subs/nearly half do not have onlinebackup 

## Comparing customers by device protection

```{r}
device_tbl <- customer_df %>% group_by(deviceprotection) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
device_tbl
```
less customers have device protection than those with

## Comparing customers by tech support

```{r}
# distribution by tech support

support_tbl <- customer_df %>% group_by(techsupport) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
support_tbl
```
more customers lack tech support

## Comparing customers by streaming tv

```{r}
# distribution by streaming tv

tv_tbl <- customer_df %>% group_by(streamingtv) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
tv_tbl
```

Customers with tv streaming service and those without are comparable.


## Comparing customers by streaming movies

```{r}
# distribution by streaming tv

movies_tbl <- customer_df %>% group_by(streamingmovies) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
movies_tbl
```
Nearly equal no of customers have movie streaming 

## Comparing customers by contract

```{r}
# ditstribution by contract

contract_tbl <- customer_df %>% group_by(contract) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
contract_tbl
```
most of the customers are on short term contract or no contract relying on month on month service subscription 

## Comparing customers by paperless billing

```{r}
# distribution by paperless billing

paperless_tbl <- customer_df %>% group_by(paperlessbilling) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
paperless_tbl
```
MOST of the customers are using paperless billing 


## Comparing customers by payment method

```{r}
payment_tbl <- customer_df %>% group_by(paymentmethod) %>% 
  summarise(count=n(),
            .groups = 'drop')
 
payment_tbl
```

Electronic check is the most used payment method among the customers 

#4.Summary of findings:

Based on the results of the analysis, the following conclusions were arrived at:

-There is no significant difference in churn rate between male and female subscribers. So this is not an area management needs to worry about.
-Majority of the customers are not senior citizens so this makes this dataset biased and hard to identify whether being a senior citizen affects churn rate.
-Not having a partner increases the likelihood of churning.
-Not having dependents increases the likelihood of churning.
-More customers with electronic check payment method are more likely to churn out - the mgt can attend more to this among other efforts 

-More churn is seen in customers with paperless billing

-More churn noted in Non partners than with partner subs

-Short term contract customers (month-month) are likely to churn out

-Churn is less/unlikely affected by availability of internet services


#5.Recommendations

In order to create an effective customer retention program, management should take the following measures:

-Focus more on meeting the needs of non-senior citizens.
-Focus more on having customers that have partners and/or dependents since these people are less likely to churn. Alternatively, management can come up with services specifically designed for customers without partners and/or dependents. This would require additional research.
-The mgt to put more efforts on why more customers who pay by electrocnic check methon churn out more 
-The mgt to come up with programs for customers to tke longer contracts to reduce the churn seen in short term contracts (mnth on month)


#6.Challenging your Solution

## a) Did we have the right data? Do we need other data to answer our question?

As far as I can tell, we had the right data. However, more data is still needed, particularly those with more customers who churned so we can have a better understanding of why they might have churned.

## b) Did we have the right question?

Yes, we did but we can add others 










```{r}

```

